The Economics of Wildfire Aviation

### 2023-09-06 manuscript/manuscript.qmd
incorporated Cal's edits on AFUE

The objective of this paper is to describe and analyze two economic aspects of aviation used for wildfire response. First, we discuss how the benefits of aviation are not well measured or tracked, making it difficulty to analyze the benefits relative to the costs. Second, we apply economic principles to the procurement of aviation resources and discuss how and where economic efficiency can be gained. While this paper will not come to clear conclusions about the use of aviation, it will frame the economic problems and attempt to provide a path forward for how to solve them.

@@

Until 2008?, federal agencies owned and operated a fleet of aircraft bearing the cost of maintenance. Since 2008 (check this), the aircraft used by federal agencies are owned by private entities. Contracts are either exclusive use (EXU) or call-when-needed (CWN) where the name references the structure of the contract. Exclusive use contracts mandate that an aircraft is exclusively available to the contracting entity (e.g., US Forest Service) for a specified number of days at an agreed upon daily availability rate and hourly flight cost. Call-when-needed contracts specify that an aircraft is available for a short period of time at a daily availability rate and hourly flight cost that generally exceeds the exclusive use rates.

@@

Aviation resources such as large airtankers (LAT) are believed to be most effective during the initial attack phase (the first burning period which is usually from ignition to 10:00 AM the following day). The logic is simple, aggressively attack the fire while it is small and it will not grow large. Most fires (more than 90%) are contained during initial attack. However, the fires that do escape initial containment because of weather or lack of sufficient IA effort can grow large and pose risk to values at risk.[^1]

@@

The fundamental question is: when is using aviation to fight fire worth the cost. The answer depends on the objectives of the mission, which aircraft are conducting the mission, and tactically where on the fire aviation is used in relation to other suppression activities. The AFUE report provides the best available information on the effectiveness of aviation resources in accomplishing stated objectives. It finds that...

Despite the advances in understanding aviation effectiveness, the AFUE report does not apply a study design capable of measuring the benefits of aviation. Describe the study design...

@@

Figure xx illustrates the empirical strategy on a retardant drop on the xx fire. The fire is progressing south and retardant is dropped in advance of the fire. The drop is xx meters long. The north side of the drop is the pre-treatment period while the south side is the post-treatment period. The control region is east of the retardant line with the north and south similarly defined as the pre and post period. In this example, the fire does not progress south of the drop into region B, but it does progress into region D (post-period control area), providing some evidence that the drop prevented the spread of the fire south.

We cannot draw conclusions from a single drop. We repeat this procedure across many drops on many fires. Once the regions are defined for each drop, we can use GIS to extract information on vegetation, topography, and weather conditions at the time the fire encountered the drop. The following model can be used to estimate the average treatment effect on the treated,

@@

The USFS determined that owning and maintaining its own fleet of aircraft is not worth the cost. In xxxx, they changed to a procurement model in which private firefighting aircraft are contracted for a period of time at a specified cost. The contacts specify a number of safety requirements, but also a daily availability price and an hourly flight price (cite Stoneseifer). The two forms of contract are Exclusive Use (EXU) and Call-When-Needed (CNW). Exclusive use contracts mandate that an aircraft is exclusively available to the contracting entity (e.g., US Forest Service) for a specified number of days at a daily cost. Call-when-needed contracts are more flexible and can be activated on short notice. However, CWN contacts generally specify a daily availability rate and hourly flight rate that is higher than EXU contracts. The question is: how many of each contract should the USFS sign at the start of each fire season?[^2] The answer depends on expected use patterns of the aviation resources.

>>>>

The objective of this paper is to describe and analyze two economic aspects of aviation used for wildfire response. First, we discuss how the benefits of aviation are not well measured or tracked, making it difficult to analyze the benefits relative to the costs. Second, we apply economic principles to the procurement of aviation resources and discuss how and where economic efficiency can be gained. While this paper will not come to clear conclusions about the use of aviation, it will frame the economic problems and attempt to provide a path forward for how to solve them.

@@

Until 2004, federal agencies owned and operated a fleet of aircraft bearing the cost of maintenance. Since 2004, federal agencies began contracting some aviation resources owned by private entities. Contracts are either exclusive use (EXU) or call-when-needed (CWN) where the name references the structure of the contract. Exclusive use contracts mandate that an aircraft is exclusively available to the contracting entity (e.g., US Forest Service) for a specified number of days at an agreed upon daily availability rate and hourly flight cost. Call-when-needed contracts specify that an aircraft is available for a short period of time at a daily availability rate and hourly flight cost that generally exceeds the exclusive use rates.

@@

Aviation resources such as large airtankers (LAT) are believed to be most effective during the initial attack phase (the first burning period which is usually from ignition to 10:00 AM the following day). The logic is simple, aggressively attack the fire while it is small and it will not grow large. Most fires (more than 90%) are contained during initial attack. However, the fires that do escape initial containment because of weather or lack of sufficient IA effort can grow large and threaten values at risk.[^1]

@@

The fundamental question is: when is using aviation to fight fire worth the cost. The answer depends on the objectives of the mission, which aircraft are conducting the mission, and tactically where on the fire aviation is used in relation to other suppression activities. The AFUE report provides the best available information on the effectiveness of aviation resources in accomplishing stated objectives. It finds that across all types of aviation, there is an 82% success rate in accomplishing mission objectives.

Despite the advances in understanding aviation effectiveness, the AFUE report does not apply a study design capable of measuring the benefits of aviation. Researchers collaborated with forest stakeholders to come up with performance metrics and evaluation criteria for analyzing aviation use on wildfires, including did retardant or water drops reduce fire intensity, delay fire growth, or extinguish the fire. Whether the objective was accomplished or not was subjectively assessed by agents in the field rather than through objective measurement by instrument. These methods are susceptible to human biases that may want to show that aviation resources provide benefits.

@@

Figure xx illustrates the empirical strategy on a retardant drop on the Terek fire that burned in Wyoming in 2018. The fire is progressing south and retardant is dropped in advance of the fire. The drop is 529 meters long. The north side of the drop is the pre-treatment period while the south side is the post-treatment period. The control region is east of the retardant line with the north and south similarly defined as the pre and post period. In this example, the fire does not progress south of the drop into region B, but it does progress into region D (post-period control area), providing some evidence that the drop prevented the spread of the fire to the south.

We cannot draw conclusions from a single drop. We repeat this procedure across many drops on many fires. Once the regions are defined for each drop, we can use geographic information systems (GIS) to extract information on vegetation, topography, and weather conditions at the time the fire encountered the drop. The following model can be used to estimate the average treatment effect on the treated,

@@

The USFS determined that owning and maintaining its own fleet of aircraft is not worth the cost. In xxxx, they changed to a procurement model in which private firefighting aircraft are contracted for a period of time at a specified cost. The contacts specify a number of safety requirements, but also a daily availability price and an hourly flight price (cite Stoneseifer). The two forms of contract are Exclusive Use (EXU) and Call-When-Needed (CWN). Exclusive use contracts mandate that an aircraft is exclusively available to the contracting entity (e.g., US Forest Service) for a specified number of days at a daily cost. Call-when-needed contracts are more flexible and can be activated on short notice. However, CWN contacts generally specify a daily availability rate and hourly flight rate that is higher than EXU contracts. The question is: how many of each contract should the USFS sign at the start of each fire season?[^2] The answer depends on expected use patterns of the aviation resources.
